# Chord Superimposition and Substitution

### Concepts and Terms {-}
- **Soloing and Comping**
  - These concepts can be used both for soloing and harmony applications.
  
- **Chord Superimposition and Chord Substitution**
  - **Superposition** refers to adding extensions to a chord by superimposing another chord from the same scale over the original. For example, playing Am7 over Dm7 adds extensions 9 and 11 to Dm7 (notes E and G).
  - **Substitution** refers to replacing the original chord by one from a different scale, with the goal of adding tension that will be resolved. For example, playing AbmMaj7 instead of G7 in the progression G7-C can be thought of as altering the G7 with b9 and #5 (notes Ab and Eb).

- **Functioning and Static Dominant Chords**
  - A **functioning dominant** is a dominant chord that resolves a 4th up (or 5th down), to its I chord. For example, if G7 is followed by a C chord, the G7 is a functioning dominant.
  - A **static dominant** is a dominant chord that does not resolve to its I. For example, if G7 is followed by E7, the G7 is a static dominant.

- **Triads and 7th Chords**
  - The concepts in this section are illustrated mostly using 7th chords. However, not every substitution has to use all notes from the new chord - the sub can be a triad, or a 2 note chord.
  

## Diatonic Superimpositions

Diatonic superimpositions involve replacing a chord with another chord from the same scale.
The replacement chords are built from degrees of the original chord, and are in the same parent scale (e.g. major). This replacement adds extensions to a chord in order of the scale degrees.

Superimpositions are an alternative way, or maybe a shorthand, of thinking about chords with extensions. Instead of thinking "Dm13" to create a minor 13 sound over Dm7, one might think "Cmaj7" and play Cmaj7 over Dm7, creating the minor 13 sound.

A superposition built on the 3rd of the original chord is also called "3-to-9", since the 3rd, 5th, 7th, and 9th of the original chord are now being played. 

**Example 1: sups over Cmaj7 using the C major scale**: **Cmaj7** has the notes C, E, G, B. The chord built on the 3rd (E) is **Em7** (notes E, G, B, D). Playing **Em7** over **Cmaj7** adds a 9th to the original **Cmaj7**, making it into **Cmaj9**. Playing **G7** (G B D F) over **Cmaj7** adds the extensions 9 and 11, and playing **Bm7b5** (B, C, F, A) adds the extensions 9, 11, and 13.

**Example 2: sups over Fmaj7 in the C major scale**: **Fmaj7** has the notes F, A, C, E. The chord built on the 3rd (A) is **Am7** (notes A, C, E, G). Playing **Am7** over **Fmaj7** adds a 9th to the original **Fmaj7**. However, playing **Cmaj7** (C E G B) over **Fmaj7** adds the notes G (9th) and B (#11), not 9th and 11. This is because the C major scale has a natural B, not a Bb, so using it for building the new chord yields the #11 extension.

In the following two tables, "Orig. Chord" is the original chord, "Subs built on chord tones" contains the diatonic chords built on the 3rd, 5th, and 7th (or flat 7th) degrees of the original chords, and "Extensions" contains all the extensions added (the sub on 7th degree provides all 3 extensions). The other set of columns provides an example in G. Concretely, for the original chord ii (Am7 in G major scale), the diatonic chord built on the 5th degree is the vi chord of the parent scale (Em7 in G major), and playing vi over ii yields extensions 9 and 11.


```{r echo=FALSE}
display_sub_table <- function(odf, caption, cnames, abv_header, font_size) {
  
  colnames(odf) <- cnames
  
  odf %>%
    kbl(caption = caption, align = "r") %>%
    kable_styling(font_size = font_size) %>%
    column_spec(1, bold = TRUE) %>%
    column_spec(1:length(cnames), border_right = T, border_left = T) %>%
    add_header_above(abv_header, background = "#e8f2a1")
}

sup_cnames <- c("Orig. Chord", "3", "5", "7 (or b7)", "Extensions","Orig. Chord", "3", "5", "7 (or b7)", "Extensions")
sup_header <- c(" " = 1, "Sups built on chord tones" = 3, " " = 1, "Example in G" = 5)


maj_scale_sups_df <- data.frame(SC = c("I", "ii", "IV", "V", "vi"),
                    sub_3 = c("iii", "IV", "vi", "vii", "I"),
                    sub_5 = c("V", "vi", "I", "ii", "iii"),
                    sub_7 = c("vii", "I", "iii", "IV", "V"),
                    exts = c("9, 11, 13", "9, 11, 13", "9, #11, 13", "9, 11, 13", "9, 11, b13"),
                    CD = c("Gmaj7", "Am7", "Cmaj7", "D7", "Em7"),
                    sub_3_ex = c("Bm7", "Cmaj7", "Em7", "F#m7b5", "Gmaj7"),
                    sub_5_ex = c("D7", "Em7", "Gmaj7", "Am7", "Bm7"),
                    sub_7_ex = c("F#m7b5", "Gmaj7", "Bm7", "Cmaj7", "D7"),
                    exts_ex = c("A, C, E", "B, D, F#", "D, F#, A", "E, G, B", "F#, A, C"))


display_sub_table(odf = maj_scale_sups_df, caption = "Diatonic Substitutions with Major Scale",
                    cnames = sup_cnames,
                    abv_header = sup_header,
                    font_size = 12)

hm_scale_sups_df <- data.frame(SC = c("imMaj7", "iim7b5", "iv", "V", "VI"),
                    sub_3 = c("III#5", "iv", "VI", "vii\u00B07", "imMaj7"),
                    sub_5 = c("V", "VI", "imMaj7", "iim7b5", "III#5"),
                    sub_7 = c("vii\u00B07", "imMaj7", "III#5", "iv", "V"),
                    exts = c("9, 11, 13", "b9, 11, 13", "9, #11, 13", "b9, 11, b13", "#9, #11, 13"),
                    CD = c("GmMaj7", "Am7b5", "Cm7", "D7", "Ebmaj7"),
                    sub_3_ex = c("Bbmaj7#5", "Cm7", "Ebmaj7", "F#dim\u00B07", "GmMaj7"),
                    sub_5_ex = c("D7", "Emaj7", "GmMaj7", "Am7b5", "Bmaj7#5"),
                    sub_7_ex = c("F#dim\u00B07", "GmMaj7", "Bbmaj7#5", "Cm7", "D7"),
                    exts_ex = c("A, C, Eb", "Bb, D, F#", "D, F#, A", "Eb, G, Bb", "F#, A, C"))

display_sub_table(odf = hm_scale_sups_df, caption = "Diatonic Substitutions with Harmonic Minor Scale",
                    cnames = sup_cnames,
                    abv_header = sup_header,
                    font_size = 12)


```

- Examples in progressions:
  - **In C major: Dm7 - G7 - C6 -> Am7 - Bm7b5 - Em7**
    - Am7 starts on the 5th of Dm7, Bm7b5 starts on the 3rd of G7, Em7 starts on the 3rd of C6.
    - The result is equivalent to **Dm11 - G9 - C9**.
  - **In C minor: Dm7b5 - G7 - Cm6 -> Fm7 - Bdim - Cm6**
    - Fm7 starts on the 3rd of Dm7b5, Bdim starts on the 3rd of G7.
    - The result is equivalent to **Dm9b5 - G9 - Cm6**.
    



## Melodic Minor for Functioning Dominants

Chords from the 7th mode of the Melodic Minor Scale (Altered Scale) can be used to add tension to a functioning dominant 7th chord.

The Altered Scale played over the V yields the following chord degrees (relative to the V): R, b9, #9, 3, b5, #5, b7. This scale thus introduces 4 new notes to the G7 chord - b9 and #9 instead of the 9th, and b5 and #5 instead of the 5th. Chords from the Altered Scale contain the alterations in various combinations. 
The notes of the Altered Scale for a dominant are the same as the notes of the Melodic Minor Scale 1/2 step up from the dominant root. E.g., the G Altered Scale is Ab Melodic Minor.

While it is possible to use alterations on a dominant resolving to a minor chord, tension in such dominant chords is created by adding notes from the Harmonic Minor or other scales (e.g. C Harmonic Minor over G7 in the progression G7-Cm).

**Example: G altered**: Playing the Ab Melodic Minor (G Altered Scale) over G7 yields the notes G, Ab, Bb, Cb, Bb, Db, Eb, F. This scale contains G, Cb, and F, which are the Root, 3rd, and b7 of G7. It also contains Ab and Bb (b9 and #9) and Db and Eb (b5 and #5). To add alterations to the harmony for G7, play chords from Ab Melodic Minor over the G7.


The following table summarizes alterations created by substituting a dominant chord with substituting with chords from the related Melodic Minor Scale. "MM Chord" is the chord in Melodic Minor, and "MM Intervals" are the MM intervals in that chord. "V7alt Intervals" are the intervals of the same notes in relation to the dominant chord. "V7alt Chord" is the new chord created by playing the MM chord over the dominant. The last 3 columns contain examples over the G7 chord.

```{r echo=FALSE}
alt_subs <- data.frame(MMC = c("imMaj7", "iim6", "iim7", "III#5", "IV7", "V7", "vim7b5", "viim7b5"),
                    MMint = c("R-b3-5-w7", "R-b3-5-6", "R-b3-5-b7", "R-3-#5-7", "R-3-5-b7", "R-3-5-b7", "R-b3-b5-b7", "R-b3-b5-b7"),
                    V7int = c("b9-3-#5-R", "b9-3-#5-b7", "#9-b5-b7-b9", "3-#5-R-#9", "b5-b7-b9-3", "#5-R-#9-b5", "b7-b9-3-#5", "R-#9-b5-b7"),
                    V7ch = c("V7b9#5", "V7b9#5", "V7b9#9b5", "V7#9#5", "V7b9b5", "V7#9b5#5", "V7b9#5", "V7#9b5"),
                    ex_MMC = c("AbmMaj7", "Abm6", "Bbm7", "BMaj7#5", "Db7", "Eb7", "Fm7b5", "Gm7b5"),
                    ex_notes = c("Ab-C-Eb-G", "Ab-C-Eb-F", "Bb-Db-F-Ab", "B-D#-G-A#", "Db-F-Ab-B", "Eb-G-Bb-Db", "F-Ab-B-Eb", "G-Bb-Db-F"),
                    ex_V7ch = c("G7b9#5", "G7b9#5", "G7b9#9b5", "G7#9#5", "G7b9b5", "G7#9b5#5", "G7b9#5", "G7#9b5")
)


display_sub_table(odf = alt_subs, caption = "Functioning Dominant Substitutions with Melodic Minor Scale",
                    cnames = c("MM Chord", "MM Intervals", "V7alt Intervals", "V7alt Chord", "MM Chord", "Notes", "G7alt Chord"),
                    abv_header = c(" " = 4, "Example in G" = 3),
                    font_size = 12)

```

- Examples in progressions:
  - **In C major: Dm7 - G7 - C6 -> Dm7 - Abm6 - C6**
    - Abm6 contains b9 and #5.
  - **In Bb major: F7 - Bbmaj7 -> D7 - Bbmaj7**
    - F7 contains #9, b5, and #5.
    

## Tritone substitution

## Dominants b3 away
A functioning dominant can be substituted by any other dominant that is located a multiple of minor 3rd away. Barry Harris refers to this as "playing with your sisters and brothers". Such substitutions yield interesting note choices, and are a different approach to alterations than using the Melodic Minor Scale.

## Adding movement to a chord
- **Secondary dominants and ii chords**
  - Any 

- Barry Harris diminished approach


## References
- [Superposition vs Substitution, with example over F blues (ps://www.jamieholroydguitar.com)](https://www.jamieholroydguitar.com/chord-substitution-vs-chord-superimposition/)
- [Functioning & Static Dominants (justinguitar.com)](https://www.justinguitar.com/guitar-lessons/functioning-static-dominants-ja-024)