# Chord Substitution

This chapter attempts to organize information that has to do with replacing chords with others. There are many different ways of thinking about this, and what follows describes just some of them.

## Concepts and Terms {-}
- **Soloing and Comping**
  - The concepts described here are used both for soloing and comping/harmony.

- **Triads and 7th Chords**
  - Most ideas here are illustrated using 7th chords. However, not every substitution has to use all notes from the new chord - the sub can be a triad, or a 2 note chord.

- **Chord Superimposition and Chord Substitution**
  - **superimposition** refers to adding extensions to a chord by superimposing another chord from the same scale over the original. For example, playing Am7 over Dm7 adds extensions 9 and 11 to Dm7 (notes E and G).

  - **Substitution** refers to replacing the original chord by one from a different scale, frequently with the goal of adding tension that will be resolved. For example, playing AbmMaj7 instead of G7 in the progression G7-C can be thought of as altering the G7 with b9 and #5 (notes Ab and Eb).

- **Functioning and Static Dominant Chords**
  - A **functioning dominant** is a dominant chord that resolves a 4th up (or 5th down), to its I (or i) chord. For example, if G7 is followed by a C chord, the G7 is a functioning dominant.
  - A **static dominant** is a dominant chord that does not resolve in the same way. For example, if G7 is followed by F7, the G7 is a static dominant.

- **Alterations and Extensions**
  - A chord is altered when one or more of its notes is raised or lowered 1/2 step. These new notes are not in the chord's diatonic scale.
  - A chord is extended when notes are added to it from its diatonic scale.

- **Altered Dominants**
  - An altered dominant 7th chord contains the R, 3, and b7 notes, and one or more altered notes (notes that are raised or lowered by 1/2 step): b9, #9, b5, #5.

- **Dominant Resolving to Major or Minor**
  - In major V7-I progressions, alterations on the dominant come from outside the parent Major Scale (e.g. using the Altered Scale). In minor V7-i, alterations on the dominant come from its parent Harmonic Minor Scale, and are actually extensions. While it is not illegal to use the Altered Scale to generate alterations in a a minor V-i, it is much more common to use Harmonic Minor. This is because in a minor V7-i in jazz, the V7 harmony usually comes from Harmonic Minor (and not Natural Minor, where the V would not be a dominant chord).  


## Diatonic Superimpositions {-}

Diatonic superimpositions involve replacing a chord with another chord from the same scale.
The replacement chords are built from degrees of the original chord, and are in the same parent scale (e.g. Major Scale). This replacement adds extensions to a chord in order of the scale degrees.

Superimpositions are an alternative way, or maybe a shorthand, of thinking about chords with extensions. Instead of thinking "Dm13" to create a minor 13 sound over Dm7, one might think "Cmaj7" and play Cmaj7 over Dm7, creating the minor 13 sound.

A superimposition built on the 3rd of the original chord is also called "3-to-9", since the 3rd, 5th, 7th, and 9th of the original chord are now being played.

It seems to be common in Jazz to omit the natural 11th in major and dominant chords because it conflicts with the major 3rd. E.g., a D13 would actually be played as D7 with the 9th and 13th.

- **Example 1: superimpositions over Cmaj7 in C Major**
  - **Cmaj7** has the notes C, E, G, B. The chord built on the 3rd (E) is **Em7** (notes E, G, B, D). Playing **Em7** over **Cmaj7** adds a 9th to the original **Cmaj7**, making it into **Cmaj9**. Playing **G7** (G B D F) over **Cmaj7** adds the extensions 9 and 11, and playing **Bm7b5** (B, C, F, A) adds the extensions 9, 11, and 13.

- **Example 2: superimpositions over Fmaj7 in C Major**
  - **Fmaj7** has the notes F, A, C, E. The chord built on the 3rd (A) is **Am7** (notes A, C, E, G). Playing **Am7** over **Fmaj7** adds a 9th to the original **Fmaj7**. However, playing **Cmaj7** (C E G B) over **Fmaj7** adds the notes G (9th) and B (#11), not 9th and 11. This is because the C Major Scale has a natural B, not a Bb, so using it for building the new chord yields the #11 extension.

In the following two tables, "Orig. Chord" is the original chord, "Sups built on chord tones" contains the diatonic chords built on the 3rd, 5th, and 7th (or flat 7th) degrees of the original chords, and "Extensions" contains all the extensions added (the sup on 7th degree contains all 3 extensions). The other set of columns provides an example in G. Concretely, for the original chord ii (Am7 in G major scale), the diatonic chord built on the 5th degree is the vi chord of the parent scale (Em7 in G major), and playing vi over ii yields extensions 9 and 11.

```{r echo=FALSE}
display_sub_table <- function(odf, caption, cnames, abv_header, font_size) {
  colnames(odf) <- cnames
  odf %>%
    kbl(caption = caption, align = "r") %>%
    kable_styling(font_size = font_size) %>%
    column_spec(1, bold = TRUE) %>%
    column_spec(1:length(cnames), border_right = T, border_left = T) %>%
    add_header_above(abv_header, background = "#e8f2a1")
}

sup_cnames <- c("Orig. Chord", "3", "5", "7 (or b7)", "Extensions","Orig. Chord", "3", "5", "7 (or b7)", "Extensions")
sup_header <- c(" " = 1, "Sups built on chord tones" = 3, " " = 1, "Example in G" = 5)

maj_scale_sups_df <- data.frame(SC = c("I", "ii", "IV", "V", "vi"),
                    sub_3 = c("iii", "IV", "vi", "vii", "I"),
                    sub_5 = c("V", "vi", "I", "ii", "iii"),
                    sub_7 = c("vii", "I", "iii", "IV", "V"),
                    exts = c("9, 11, 13", "9, 11, 13", "9, #11, 13", "9, 11, 13", "9, 11, b13"),
                    CD = c("Gmaj7", "Am7", "Cmaj7", "D7", "Em7"),
                    sub_3_ex = c("Bm7", "Cmaj7", "Em7", "F#m7b5", "Gmaj7"),
                    sub_5_ex = c("D7", "Em7", "Gmaj7", "Am7", "Bm7"),
                    sub_7_ex = c("F#m7b5", "Gmaj7", "Bm7", "Cmaj7", "D7"),
                    exts_ex = c("A, C, E", "B, D, F#", "D, F#, A", "E, G, B", "F#, A, C"))

display_sub_table(odf = maj_scale_sups_df, caption = "Diatonic Substitutions with Major Scale",
                    cnames = sup_cnames,
                    abv_header = sup_header,
                    font_size = 12)

hm_scale_sups_df <- data.frame(SC = c("imMaj7", "iim7b5", "iv", "V", "VI"),
                    sub_3 = c("III#5", "iv", "VI", "vii\u00B07", "imMaj7"),
                    sub_5 = c("V", "VI", "imMaj7", "iim7b5", "III#5"),
                    sub_7 = c("vii\u00B07", "imMaj7", "III#5", "iv", "V"),
                    exts = c("9, 11, 13", "b9, 11, 13", "9, #11, 13", "b9, 11, b13", "#9, #11, 13"),
                    CD = c("GmMaj7", "Am7b5", "Cm7", "D7", "Ebmaj7"),
                    sub_3_ex = c("Bbmaj7#5", "Cm7", "Ebmaj7", "F#dim\u00B07", "GmMaj7"),
                    sub_5_ex = c("D7", "Emaj7", "GmMaj7", "Am7b5", "Bmaj7#5"),
                    sub_7_ex = c("F#dim\u00B07", "GmMaj7", "Bbmaj7#5", "Cm7", "D7"),
                    exts_ex = c("A, C, Eb", "Bb, D, F#", "D, F#, A", "Eb, G, Bb", "F#, A, C"))

display_sub_table(odf = hm_scale_sups_df, caption = "Diatonic Substitutions with Harmonic Minor Scale",
                    cnames = sup_cnames,
                    abv_header = sup_header,
                    font_size = 12)
```

- Examples in Progressions
  - **In C major: Dm7 - G7 - C6 -> Am7 - Bm7b5 - Em7**
    - **Am7** starts on the 5th of **Dm7**, **Bm7b5** starts on the 3rd of **G7**, **Em7** starts on the 3rd of **C6**.
    - The result is equivalent to **Dm11 - G9 - C9**.
  - **In C minor: Dm7b5 - G7 - Cm6 -> Fm7 - Bdim - Cm6**
    - **Fm7** starts on the 3rd of **Dm7b5**, **Bdim** starts on the 3rd of **G7**.
    - The result is equivalent to **Dm9b5 - G9 - Cm6**.

## Functioning Dominant Substitutions {-}

In Jazz, functioning dominant 7th chords are frequently played differently than written on a lead sheet,  to add tension and interest.

Using alterations (described in the **Altered Scale** section) is a standard way to add tension, and one way of remembering these alterations is to simply memorize new voicings that contain them. Another way is to use chord substitutions - similarly to supersimpositions, we can use familiar shapes to generate new voicings.

### Altered Scale {-}

Chords from the Altered Scale (7th mode of the Melodic Minor Scale) can be used to alter a functioning dominant 7th chord in a major V7-I progression.

The Altered Scale played over its related dominant chord yields the following chord degrees relative to the dominant: R, b9, #9, 3, b5, #5, b7. This scale thus alters the 9th and the 5th of the chord - b9 and #9 instead of 9, and b5 and #5 instead of 5. 

The notes of the Altered Scale for a dominant are the same as the notes of the Melodic Minor Scale 1/2 step up from the dominant root (the Altered Scale is equivalent to the 7th mode of Melodic Minor). E.g., the G Altered Scale has the same notes as Ab Melodic Minor.

Chords constructed using the Altered Scale contain the alterations in various combinations.

- **Example: G7alt**
  - Playing the Ab Melodic Minor (G Altered Scale) over **G7** yields the notes **G, Ab, Bb, Cb, Bb, Db, Eb, F**. This scale contains **G**, **Cb**, and **F**, which are the Root, 3rd, and b7 of **G7**. It also contains **Ab** and **Bb** (b9 and #9) and **Db** and **Eb** (b5 and #5). To use alterations over the harmony for **G7**, play chords from Ab Melodic Minor over the **G7**.

The following table summarizes alterations created by substituting a dominant chord with chords from the related Melodic Minor Scale. "MM Chord" is the chord in Melodic Minor, and "MM Intervals" are the MM intervals in that chord. "V7alt Intervals" are the intervals of the same notes in relation to the dominant chord. "V7alt Chord" is the new chord created by playing the MM chord over the dominant. The last 3 columns contain examples over a G7 chord.

```{r echo=FALSE}
alt_subs <- data.frame(MMC = c("imMaj7", "iim6", "iim7", "III#5", "IV7", "V7", "vim7b5", "viim7b5"),
                    MMint = c("R-b3-5-7", "R-b3-5-6", "R-b3-5-b7", "R-3-#5-7", "R-3-5-b7", "R-3-5-b7", "R-b3-b5-b7", "R-b3-b5-b7"),
                    V7int = c("b9-3-#5-R", "b9-3-#5-b7", "#9-b5-b7-b9", "3-#5-R-#9", "b5-b7-b9-3", "#5-R-#9-b5", "b7-b9-3-#5", "R-#9-b5-b7"),
                    V7ch = c("V7b9#5", "V7b9#5", "V7b9#9b5", "V7#9#5", "V7b9b5", "V7#9b5#5", "V7b9#5", "V7#9b5"),
                    ex_MMC = c("AbmMaj7", "Abm6", "Bbm7", "BMaj7#5", "Db7", "Eb7", "Fm7b5", "Gm7b5"),
                    ex_notes = c("Ab-C-Eb-G", "Ab-C-Eb-F", "Bb-Db-F-Ab", "B-D#-G-A#", "Db-F-Ab-B", "Eb-G-Bb-Db", "F-Ab-B-Eb", "G-Bb-Db-F"),
                    ex_V7ch = c("G7b9#5", "G7b9#5", "G7b9#9b5", "G7#9#5", "G7b9b5", "G7#9b5#5", "G7b9#5", "G7#9b5")
)


display_sub_table(odf = alt_subs, caption = "Functioning Dominant Substitutions with Melodic Minor Scale",
                    cnames = c("MM Chord", "MM Intervals", "V7alt Intervals", "V7alt Chord", "MM Chord", "Notes", "G7alt Chord"),
                    abv_header = c(" " = 4, "Example in G" = 3),
                    font_size = 12)

```

- **Examples in Progressions**
  - **In C major: Dm7 - G7 - C6 -> Dm7 - Abm6 - C6**
    - **Abm6** contains b9 and #5.
  - **In Bb major: F7 - Bbmaj7 -> D7 - Bbmaj7**
    - **D7** contains #9, b5, and #5.
    

### Tritone {-}

- **Examples in Progressions**
  - **A7 - Dm7 - G7 - C -> Eb7 - Dm7 - Db7 - C**
    - The **A7** and **G7** are replaced with dominants a tritone away, creating a descending chromatic progression.

### Dominants b3 Away {-}
This idea is based on Barry Harris's approach to using related dominant scales (located in increments of minor 3rd away) over a V chord (referred to as "playing with your sisters and brothers"). Applying this to chords, a functioning dominant can be substituted by any other dominant that is located a multiple of a minor 3rd away.

One of the substituting dominant chords (b5 away) is the tritone sub.

- **Example 1: G7**
  - Building a dom7 chord on the bb7 of **G7** (3 multiples of b3 up) creates **E7**, which contains the notes **E**, **Ab**, **B**, **D**. These are 6, b9, 3, and 5 in reference to **G7** - so substituting **E7** creates a **G7** chord with with 13 and b9 notes.

In the following table, "Sub Chord" is the substitute chord (with the first row simply being the original chord), "Orig. Chord Degrees" are the degrees of the sub in relation to the original chord, and the last two columns show examples over G7.

```{r echo=FALSE}
d7_b3_sub_df <- data.frame(TP = c("dom7 on R","dom7 on b3","dom7 on b5","dom7 on bb7"),
                           D7F = c("R-3-5-b7","#9-5-b7-b9","b5-b7-b9-3","6-b9-3-5"),
                           ex_d7 = c("G7","Bb7","Db7","E7"),
                           ex_notes = c("G-B-D-F","Bb-D-F-Ab","Db-F-Ab-B","E-Ab-B-D"))

display_sub_table(odf = d7_b3_sub_df, caption = "Functioning Dominant Substitutions b3 Away",
                    cnames = c("Sub Chord", "Orig. Chord Degrees", "Sub Chord", "Notes"),
                    abv_header = c(" " = 2, "Example in G" = 2),
                    font_size = 12)

```

- **Examples in Progressions**
  - **A7 - Dm7 - G7 - C -> C7 - Dm7 - Bb7 - C**
    - **A7** is replaced with **C7** (built on bb7), and **G7** is replaced with **Bb7** (built on b3).




## Adding movement to a chord {-}
This section describes a different approach to substitutions - instead of replacing a chord with another chord, insert other chords before the original chord, breaking up how long the original chord lasts.

### Secondary dominants and ii chords {-}
  - Major, minor, and dominant chords (not m7b5) can be preceded by a secondary dominant - a dominant chord that resolves a 4th up (or 5th down). The secondary dominants (besides the main V) are not in the original scale, but can be played before their resolution chords to add interest. For example, the V chord in C Major is G7 (the dominant of Cmaj7); Dm7's secondary dominant is A7, Em7's is B7, and so on.
  - Taking this further, a functioning dominant can be preceeded by its ii chord. This can also be combined with a tritone sub on the dominant - the ii is now relative to the tritone. Generally, the ii chord is minor 7 when resolving to major or maj7, and m7b5 when resolving to a minor, m7 or m6 - but this is not always the case.

- **Examples in Progressions**
  - **Am7 - Cmaj7 -> E7 - Am7 - Dm7 - G7 - Cmaj7**
    - **E7** is the secondary dominant of **Am7**. **G7** is the secondary dominant of **Cmaj7**, and **Dm7** is the ii of **G7**.
  - **G7 - C -> Abm7 - Db7 - C**
    - **Db7** is the tritone sub of **G7**, and **Abm7** is the ii of **Db7**.
  - **Cm -> Dm7b5 - G7 - Cm**
    - **G7** is the secondary dominant of **Cm**, and **Dm7b5** is the iim7b5 of **G7**.

- **Barry Harris diminished approach**


## References {-}
- [superimposition vs Substitution, with example over F blues (ps://www.jamieholroydguitar.com)](https://www.jamieholroydguitar.com/chord-substitution-vs-chord-superimposition/)
- [Functioning & Static Dominants (justinguitar.com)](https://www.justinguitar.com/guitar-lessons/functioning-static-dominants-ja-024)
- [Everything You Don’t Know About Minor Harmony in Jazz](https://www.jazzadvice.com/lessons/minor-keys-and-harmony-in-jazz/) - An excellent detailed explanation of how dominant chords work in minor-key progressions.
- [Altered Chord (Wikipedia)](https://en.wikipedia.org/wiki/Altered_chord)